### 2014-12-07 doc/thesis/system.tex
Improvements to the text about table types

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
\multicolumn{3}{c}{\textbf{First-level types}}\\
t ::= & \;\; l & \textit{literal types}\\
& | \; b & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; t \cup t & \textit{disjoint union types}\\
& | \; p \rightarrow r & \textit{function types}\\
& | \; \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{closed|open|unique} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.t & \textit{recursive types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
l ::= & \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} & \\
b ::= & \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String & \\
k ::= & \;\; l \; | \; b \; | \; \Any & \textit{key types}\\
v ::= & \;\; t \; | \; \Const \; t & \textit{field types}\\ 
\multicolumn{3}{c}{}\\
\multicolumn{3}{c}{\textbf{Second-level types}}\\
s ::= & \;\; t &\\
& | \; t* & \textit{variadic types}\\
& | \; t \times s & \textit{tuple types}\\
p ::= & \;\; \Void \; | \; s & \textit{parameters list type}\\
r ::= & \;\; \Void \; | \; s \; | \; s \sqcup r & \textit{return types}\\
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

@@

We also use the tags \emph{closed}, \emph{open}, and \emph{unique}
to classify table types.
The first tag represents the table types that the type annotations,
interface, and userdata declarations describe.
The second tag represents the table types that we can use field assignment
to add new fields to them.
The third tag represents the table types of table constructors.
We have different subtyping rules for table types according to their
tags.
We will discuss these subtyping rules in the next paragraphs to
explain why we needed different tags.

The following subtyping rule shows that \emph{closed} table types
have width subtyping, and depth subtyping on $\Const$ fields;
it resembles the subtyping rule for record types.

\[
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
       \senv \vdash k_{j} \subtype k_{i}' \;\;\;
       \senv \vdash k_{i}' \subtype k_{j} \;\;\;
       \senv \vdash v_{j} \subtype_{c} v_{i}'}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{closed} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
\end{array}
\]

The rule \textsc{S-TABLE1} introduced the relation $\subtype_{c}$.
It is just an auxiliary relation that defines the subtyping of the
type of the values stored in table fields.
We needed an auxiliary relation because the subtyping of the type
of the values stored in the table fields change according to the
tag of the table type.
For closed tables, subtyping handles mutable fields invariantly,
and immutable fields covariantly.
The following rules define this behavior.
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2} \;\;\;
       \senv \vdash v_{2} \subtype v_{1}}
      {\senv \vdash v_{1} \subtype_{c} v_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash \Const \; v_{1} \subtype_{c} \Const \; v_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash v_{1} \subtype_{c} \Const \; v_{2}}
\end{array}
\end{array}
\]

There are two limitations on \emph{closed} table types that led us to
introduce \emph{open} and \emph{unique} table types.
First, they do not allow programmers to use the table constructor to
initialize table types that include optional fields.
For instance,
\begin{align*}
\{ & ``firstname":\String, ``lastname":\String \}_{closed} \not \subtype\\
\{ & ``firstname":\String,\\
  &  ``middlename":\String \cup \Nil, ``lastname":\String \}_{closed}
\end{align*}
Second, they do not allow programmers to use the table constructor to
initialize table types that represent hashes.
For instance,
\begin{align*}
\{ & ``x":\Number, ``y":\Number, ``z":\Number \}_{closed} \not \subtype\\
\{ & \String:\Number \cup \Nil\}_{closed}
\end{align*}

@@

Table types can be ambiguous.
For instance, $\{1:\Number, \Integer:\String\}$ is an ambiguous
table type because the type of the value stored by key $1$ can be
$\Number$ or $\String$, as $1 \subtype 1$ and
$1 \subtype \Integer$.
Table types can get even more ambiguous when we use the dynamic
type in the type of table keys.
For instance, the table type $\{1:\Number, \Integer:\String, \Any:\Boolean\}$
is ambiguous because the type of the value stored by key $1$ can be
$\Number$, or $\String$, or $\Boolean$, as $1 \subtype 1$,
$1 \subtype \Integer$, and $1 \lesssim \Any$.
Moreover, the type of the value stored by a key of type $\Integer$,
which is not the literal type $1$, can be $\Number$ or $\Boolean$,
as $\Integer \subtype \Integer$, and $\Integer \lesssim \Any$.

To avoid ambiguous table types we use the following rule to check
whether a table type is well formed:
\[
\forall i \not\exists j \; i \not= j \wedge k_{i} \lesssim k_{j}
\]

Our type system also includes recursive types: $\mu x.t$,
where $t$ is a function or table type with $x$ appearing anywhere
a function or table type could appear.
For instance, $\mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\}_{closed}$
represents the type of single linked list of numbers.

We use the \emph{Amber rule} from the Amber language
\citep{cardelli1986amber} to define the subtyping rule for
recursive types:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-ASSUMPTION}\\
\dfrac{x_{1} \subtype x_{2} \in \senv}
      {\senv \vdash x_{1} \subtype x_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-AMBER}\\
\dfrac{\senv[x_{1} \subtype x_{2}] \vdash t_{1} \subtype t_{2}}
      {\senv \vdash \mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}}
\end{array}
\end{array}
\]

Our type system does not infer recursive types, so we defined two
extra subtyping rules that handle the cases where they may
appear inside a table or function type:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-AMBERL}\\
\dfrac{\senv \vdash [x \mapsto \mu x.t_{1}]t_{1} \subtype t_{2}}
      {\senv \vdash \mu x.t_{1} \subtype t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-AMBERR}\\
\dfrac{\senv \vdash t_{1} \subtype [x \mapsto \mu x.t_{2}]t_{2}}
      {\senv \vdash t_{1} \subtype \mu x.t_{2}}
\end{array}
\end{array}
\]

For instance, the rule \textsc{S-AMBERR} allows our type system
check whether the type of a table constructor is subtype of
a recursive type or not, as we show in the following example:
\begin{align*}
\{ & ``info":\Number, \\
   & ``next":\mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\}_{closed} \cup \Nil \}_{unique} \subtype \\
& \mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\}_{closed}
\end{align*}

Our type system also introduces projection types, but they do not
appear on the subtyping and consistent-subtyping rules.
They are just a mechanism that the type system uses for handling
union of tuple types, when they appear on the right-hand side of
the declaration of local variables.
That is the reason why we do not need to define subtyping and
consistent-subtyping rules for them.
We will see projection types in more detail in the next section.

In this section we use a reduced core of Typed Lua to present the
most interesting typing rules of our type system.
It is also a reduced core of the Lua language with only if and while
control flow statements, without syntactic sugar, with explicit
type annotations, with explicit scope of local variables, and
with explicit declaration of methods.
We use this reduced core because it simplifies the presentation
of our type system.
Appendix \ref{app:rules} presents the full set of typing rules.

Our reduced code also splits function application, method application, and
the vararg expression (${...}$) into different categories.
It uses $e(el)_{s}$ to denote function applications that return no value
because they appear as statements,
$e(el)_{e}$ to denote function applications that return only one value,
and $e(el)_{m}$ to denote function applications that return multiple values.
It uses the same categories for method applications, but only ${...}_{e}$
and ${...}_{m}$, as the vararg expression cannot appear as a statement.

>>>>

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
\multicolumn{3}{c}{\textbf{First-level types}}\\
t ::= & \;\; l & \textit{literal types}\\
& | \; b & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; t \cup t & \textit{disjoint union types}\\
& | \; p \rightarrow r & \textit{function types}\\
& | \; \{k_{i}{:}v_{i}, ..., k_{n}{:}v_{n}\}_{closed|open|unique} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.t & \textit{recursive types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
l ::= & \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} & \\
b ::= & \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String & \\
k ::= & \;\; l \; | \; b \; | \; \Any & \textit{key types}\\
v ::= & \;\; t \; | \; \Const \; t & \textit{value types}\\ 
\multicolumn{3}{c}{}\\
\multicolumn{3}{c}{\textbf{Second-level types}}\\
s ::= & \;\; t &\\
& | \; t* & \textit{variadic types}\\
& | \; t \times s & \textit{tuple types}\\
p ::= & \;\; \Void \; | \; s & \textit{parameters list type}\\
r ::= & \;\; \Void \; | \; s \; | \; s \sqcup r & \textit{return types}\\
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

@@

We also use the tags \emph{closed}, \emph{open}, and \emph{unique}
to classify table types.
The first tag represents the table types that the type annotations,
interface, and userdata declarations describe.
The second tag represents the table types that we can use field assignment
to add new fields to them.
The third tag represents the table types of table constructors.
We have different subtyping rules for table types according to their
tags.

The subtyping rule between \emph{closed} table types resembles the
standard subtyping rule between records:
\[
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
       \senv \vdash k_{j} \subtype k_{i}' \;\;\;
       \senv \vdash k_{i}' \subtype k_{j} \;\;\;
       \senv \vdash v_{j} \subtype v_{i}'}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{closed} \subtype \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
\end{array}
\]

The rule \textsc{S-TABLE1} allows \emph{closed} table types to
have width subtyping, and depth subtyping on $\Const$ fields.
The subtyping rules for the type of the values stored in a \emph{closed}
table type describe this behavior:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2} \;\;\;
       \senv \vdash v_{2} \subtype v_{1}}
      {\senv \vdash v_{1} \subtype v_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash \Const \; v_{1} \subtype \Const \; v_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash v_{1} \subtype \Const \; v_{2}}
\end{array}
\end{array}
\]

The rule \textsc{S-FIELD1} handles mutable values invariantly,
while the rule \textsc{S-FIELD2} handles immutable values covariantly.
It is also safe to promote a mutable value to immutable, and the rule
\textsc{S-FIELD3} defines this behavior.
However, it is not safe to promote an immutable value to mutable,
and that is the reason why we do not have a subtying rule for this case.

There is a limitation on \emph{closed} table types that led us to
introduce \emph{unique} table types.
If the table constructor had a \emph{closed} table type, then
programmers would not be able to use it to initialize a variable with
a table type that describe a more general type.
For instance,
\begin{verbatim}
    local t:{"x":integer, "y":integer?} = { x = 1, y = 2 }
\end{verbatim}
whould not type check, as the type of the table constructor would not
be subtype of the type in the annotation.
More precisely,
\[
\{``x":1, ``y":2\}_{closed} \not\subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\]
Thus, we introduced \emph{unique} table types to represent the type of
the table constructor and avoid this limitation.

We introduced \emph{open} table types to allow the refinement of table types,
an idiom very common in Lua.

Even though \emph{open} and \emph{unique} table types have specific
subtyping rules, we will present them in the next section along with
the typing rules that formalize the refinement of table types.
The presentation of these subtyping rules is easier in the next section,
as they include some unsafe behaviors that the refinement of
table types keeps track to make them safe.

Any table type has to be \emph{well-formed}.
Informally, a table type is well-formed if key types do not overlap.
Formally, a table type is well-formed if it obeys the following rule:
\[
\forall i \not\exists j \; i \not= j \wedge k_{i} \lesssim k_{j}
\]

Well-formed table types avoid ambiguous table types.
For instance, this rule detects that the table type
$\{1:\Number, \Integer:\String, \Any:\Boolean\}$ is ambiguous,
because the type of the value stored by key $1$ can be
$\Number$, $\String$, or $\Boolean$, as $1 \lesssim 1$,
$1 \lesssim \Integer$, and $1 \lesssim \Any$.
Moreover, the type of the value stored by a key of type $\Integer$,
which is not the literal type $1$, can be $\Number$ or $\Boolean$,
as $\Integer \lesssim \Integer$, and $\Integer \lesssim \Any$.

Recursive types have the form $\mu x.t$,
where $t$ is a first-level type with $x$ appearing anywhere this
first-level type could appear.
For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}$
is a type for singly-linked list of integers.

We use the \emph{Amber rule} \citep{cardelli1986amber} to define
subtyping between recursive types:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-ASSUMPTION}\\
\dfrac{x_{1} \subtype x_{2} \in \senv}
      {\senv \vdash x_{1} \subtype x_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-AMBER}\\
\dfrac{\senv[x_{1} \subtype x_{2}] \vdash t_{1} \subtype t_{2}}
      {\senv \vdash \mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}}
\end{array}
\end{array}
\]

A recursive type may appear inside a first-level type, and our
type system includes subtyping rules to handle subtyping between
a first-level type and a recursive type:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-AMBERL}\\
\dfrac{\senv \vdash [x \mapsto \mu x.t_{1}]t_{1} \subtype t_{2}}
      {\senv \vdash \mu x.t_{1} \subtype t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-AMBERR}\\
\dfrac{\senv \vdash t_{1} \subtype [x \mapsto \mu x.t_{2}]t_{2}}
      {\senv \vdash t_{1} \subtype \mu x.t_{2}}
\end{array}
\end{array}
\]

For instance, it is the rule \textsc{S-AMBERR} that allows our
type system to type check the example that we presented
in Section \ref{sec:tables}:
\begin{verbatim}
    local interface Element
      info:integer
      next:Element?
    end
    local function insert (e:Element?, v:integer):Element
      return { info = v, next = e }
    end
\end{verbatim}

The interface \texttt{Element} defines the type for singly-linked
list of integers that we just mentioned, and the function
\texttt{insert} inserts an element in the begining of the list.
In this example, the recursive type appears inside the type of
the table constructor.
It type checks because the rule \textsc{S-AMBERR} verifies that
the type of the table constructor is a subtype of \texttt{Element}:
\begin{align*}
\{ & ``info":\Integer, \\
   & ``next":\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed} \cup \Nil \}_{unique} \subtype \\
& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
\end{align*}

Projection types have the form $\pi_{i}^{x}$, where $x$ is a variable
that stores an union of tuple types in a special type environment,
and $i$ is an index that indicates how to build a first-level
union type from $x$.
For instance, if $x$ is the union of tuple types
$(\Integer \times \Integer) \sqcup (\Nil \times \String)$,
then $\pi_{1}^{x}$ represents the type $\Integer \cup \Nil$
while $\pi_{2}^{x}$ represents the type $\Integer \cup \String$.
Projection types follow the reflexive subtyping rule.
In the next section we will show in more detail how our type system
uses them as a mechanism for handling unions of tuple types,
when they appear on the right-hand side of the declaration of local variables.
We will also show how this feature allows our type system to constrain
the type of a local variable that depends on the type of another local variable.

\textit{/* How should we control invalid indices? */}

In this section we use a reduced core of Typed Lua to present the
most interesting rules of our type system.
This core limits control flow to if and while statements,
has explicit type annotations, and explicit scope for variables.
It also has explicit method declarations and explicit method calls
instead of treating them as syntactic sugar.
We use this reduced core because it simplifies the presentation
of our type system.
Appendix \ref{app:rules} presents the full set of typing rules.

Our reduced core also splits function application, method application, and
the vararg expression (${...}$) into different syntactic categories.
It uses $e(el)_{s}$ to denote function applications that return no value
because they appear as statements,
$e(el)_{e}$ to denote function applications that return only one value,
and $e(el)_{m}$ to denote function applications that return multiple values.
It uses the same categories for method applications, but only ${...}_{e}$
and ${...}_{m}$, as the vararg expression cannot appear as a statement.

@@
